\section{What is $0^0$}

  According to some Calculus textbooks, $0^0$ is an ``indeterminate
    form''. When evaluating a limit of the form $0^0$, then you need
    to know that limits of that form are called ``indeterminate forms'',
    and that you need to use a special technique such as L'Hopital's
    rule to evaluate them. Otherwise, $0^0=1$ seems to be the most
    useful choice for $0^0$. This convention allows us to extend
    definitions in different areas of mathematics that otherwise would
    require treating 0 as a special case. Notice that $0^0$ is a
    discontinuity of the function $x^y$. More importantly, keep
    in mind that the value of a function and its limit need not be the
    same thing, and functions need not be continous, if that serves
    a purpose (see Dirac's delta).


    This means that depending on the context where $0^0$ occurs, you
    might wish to substitute it with 1, indeterminate or
    undefined/nonexistent.

    Some people feel that giving a value to a function with an
    essential discontinuity at a point, such as $x^y$ at $(0,0)$, is
    an inelegant patch and should not be done. Others point out
    correctly that in mathematics, usefulness and consistency are
    very important, and that under these parameters $0^0=1$ is
    the natural choice.

    The following is a list of reasons why $0^0$ should be 1.

    Rotando \& Korn show that if $f$ and $g$ are real functions that vanish
    at the origin and are analytic at 0 (infinitely differentiable is
    not sufficient), then $f(x)^{g(x)}$ approaches 1 as $x$ approaches 0 from
    the right.

    From Concrete Mathematics p.162 (R. Graham, D. Knuth, O. Patashnik):
\begin{quote}
    Some textbooks leave the quantity $0^0$ undefined, because the
    functions $x^0$ and $0^x$ have different limiting values when $x$
    decreases to 0. But this is a mistake. We must define
       $x^0 = 1$ for all $x$,
    if the binomial theorem is to be valid when $x=0$, $y=0$, and/or $x=-y$.
    The theorem is too important to be arbitrarily restricted! By
    contrast, the function $0^x$ is quite unimportant.
\end{quote}
   Published by Addison-Wesley, 2nd printing Dec, 1988.

    As a rule of thumb, one can say that $0^0 = 1$, but $0.0^{0.0}$ is
    undefined, meaning that when approaching from a different
    direction there is no clearly predetermined value to assign to
    $0.0^{0.0}$; but Kahan has argued that $0.0^{0.0}$ should be 1,
    because if $f(x), g(x) \rightarrow 0$ as $x$ approaches some
    limit, and $f(x)$ and $g(x)$ are analytic functions, then
    $f(x)^g(x) \rightarrow 1$.


    The discussion on $0^0$ is very old, Euler argues for $0^0=1$
    since $a^0=1$ for $a \neq 0$. The controversy raged throughout the
    nineteenth century, but was mainly conducted in the pages of the
% correct name of Schlomiclh's by Edgar Fuss April, 1997
    lesser journals: Grunert's Archiv and Schlomilch's Zeitschrift f\"ur 
    Mathematik und Physik. Consensus has recently been built around 
    setting the value of $0^0=1$.

    On a discussion of the use of the function $0^{0^x}$ by an Italian
    mathematician named Guglielmo Libri.

\begin{quote}
    [T]he paper [33] did produce several ripples in mathematical
waters when
    it originally appeared, because it stirred up a controversy about whether
    $0^0$ is defined.  Most mathematicians agreed that $0^0 = 1$, but Cauchy 
[5, page 70] had listed $0^0$ together with other expressions like $0/0$ and
    $\infty-\infty$ in a table of undefined forms.  Libri's
justification for
    the equation $0^0 = 1$ was far from convincing, and a commentator
who signed
    his name simply ``S'' rose to the attack [45].  August M\"obius
[36]
    defended Libri, by presenting his former professor's reason for
believing
    that $0^0 = 1$ (basically a proof that $\lim_{x\rightarrow 0+} x^x
= 1$).  M\"obius also
    went further and presented a supposed proof that
$\lim_{x\rightarrow 0+} f(x)^{g(x)}$
    whenever $\lim_{x\rightarrow 0+} f(x) = lim_{x\rightarrow 0+} g(x)
= 0$.  Of
 course ``S'' then
    asked [3] whether M\"obius knew about functions such as $f(x) = e^{-1/x}$
    and $g(x) = x$.  (And paper [36] was quietly omitted from the historical
    record when the collected words of M\"obius were ultimately published.)
    The debate stopped there, apparently with the conclusion that $0^0$ should
    be undefined.

    But no, no, ten thousand times no!  Anybody who wants the binomial theorem
            $ (x+y)^n = \sum_{k=0}^n {n\choose k} x^k y^{n-k}$
    to hold for at least one nonnegative integer $n$ {\it must} believe that
    $0^0 = 1$, for we can plug in $x = 0$ and $y = 1$ to get 1 on the left and
$0^0$ on
    the right.

    The number of mappings from the empty set to the empty set is $0^0$.
    It {\it has} to be 1.

    On the other hand, Cauchy had good reason to consider $0^0$ as an undefined
    {\it limiting form}, in the sense that the limiting value of $f(x)^{g(x)}$ is
    not known {\it a priori} when $f(x)$ and $g(x)$ approach 0 independently.  In
    this much stronger sense, the value of $0^0$ is less defined than, say, the
    value of $0+0$.  Both Cauchy and Libri were right, but Libri and his
    defenders did not understand why truth was on their side.

    [3] \article{Anonymous and S$\ldots$}{Bemerkungen zu den Aufsatze 
\"uberschrieben,
    `Beweis der Gleichung $0^0 = 1$, nach J. F. Pfaff',}{im zweiten
Hefte dieses
    Bandes, S. 134, Journal f\"ur die reine und angewandte Mathematik,}{12
    (1834), 292--294.}

    [5] \book{\OE uvres Compl\`etes.}{Augustin-Louis Cauchy.}{Cours
d'Analyse de
    l'Ecole Royale Polytechnique
    (1821).  Series 2, volume 3.}

    [33]  \article{Guillaume Libri.}{M\'emoire sur les fonctions discontinues, 
Journal
    f\"ur die reine und angewandte Mathematik,}{}{10 (1833), 303--316.}

    [36] \article{A. F. M\"obius.}{Beweis der Gleichung $0^0 = 1$, nach J. F.
Pfaff.}
    {Journal f\"ur die reine und angewandte Mathematik,}{}{12 (1834),
134--136.}

    [45] \article{S$\ldots$}{Sur la valeur de $0^0$.}{Journal f\"ur die reine
und  angewandte Mathematik 11,}{(1834), 272--273.}

\end{quote}

    \Ref

    \article{Knuth.}{Two notes on notation.}{(AMM 99 no. 5 (May 1992),}
    {403--422).}

    \article{H. E. Vaughan.}{The expression '$0^0$'.}{Mathematics
Teacher 63 (1970),}
    {pp.111-112.}

% John W. Loux
% and  Bill Dubuque <wgd@martigny.ai.mit.edu>  

    \article{Kahan, W.} {Branch Cuts for Complex Elementary Functions
                 or Much Ado about Nothing's Sign Bit,}{The State of the Art in
                 Numerical Analysis, editors A. Iserles and M. J. D. Powell, Clarendon
                 Press, Oxford, pp. 165--212. }
    \article{Louis M. Rotando and Henry Korn.}{The Indeterminate Form $0^0$.}
    {Mathematics Magazine,}{Vol. 50, No. 1 (January 1977), pp. 41-42.}

    \article{L. J. Paige,.}{A note on indeterminate forms.}{American
Mathematical
    Monthly,}{61 (1954), 189-190; reprinted in the Mathematical
    Association of America's 1969 volume, Selected Papers on Calculus,
    pp. 210-211.}

   \article{Baxley \& Hayashi.}{A note on indeterminate
forms.}{American Mathematical
    Monthly,}{85 (1978), pp. 484-486.}


    \article{\ }{Crimes and Misdemeanors in the Computer
                 Algebra Trade.} {Notices of the American
                 Mathematical Society,}{  September 1991, volume 38, number 7,
           pp.778-785}
