\section{The Axiom of Choice}

There are several equivalent formulations:

\begin{itemize}

  \item The Cartesian product of nonempty sets is nonempty, even if the
  product is of an infinite family of sets.

  \item Given any set $S$ of mutually disjoint nonempty sets, there is a
  set $C$ containing a single member from each element of $S$.  $C$ can
  thus be thought of as the result of ``choosing'' a representative from
  each set in $S$. Hence the name.
\end{itemize}

\subsection{Relevance of the Axiom of Choice}
THE AXIOM OF CHOICE

There are many equivalent statements of the Axiom of Choice.  The
following version gave rise to its name:
\begin{quote}
  For any set $X$ there is a function $f$, with domain $X\backslash{0}$,
  so that $f(x)$ is a member of $x$ for every nonempty $x$ in $X$.
\end{quote}

Such an $f$ is called a ``choice function'' on $X$.  [Note that
$X\backslash {0}$ means $X$ with the empty set removed.  Also note that
in Zermelo-Fraenkel set theory all mathematical objects are sets so each
member of $X$ is itself a set.]

The Axiom of Choice (AC) is one of the most discussed axioms of
mathematics, perhaps second only to Euclid's parallel postulate.  The
axioms of set theory provide a foundation for modern mathematics in the
same way that Euclid's five postulates provided a foundation for
Euclidean geometry, and the questions surrounding AC are the same as the
questions that surrounded Euclid's Parallel Postulate:

\begin{enumerate}
  \item Can it be derived from the other axioms?
  \item Is it consistent with the other axioms?
  \item Should we accept it as an axiom?
\end{enumerate}

For many sets, including any finite set, the first six axioms of set
theory (abbreviated ZF) are enough to guarantee the existence of a
choice function but there do exist sets for which AC is {\em required}
to show the existence of a choice function.  The existence of such sets
was proved in 1963 by Paul Cohen.  This means that AC cannot be derived
from the other six axioms; in other words ``AC is independent of ZF.''
This answers question [1] posed above.

The question of whether AC is consistent with the other axioms (question
[2] above) was answered by Goedel in 1938.  Goedel showed that if the
other axioms are consistent then AC is consistent with them.  This is a
``relative consistency'' proof which is the best we can hope for because
of Goedel's Second Incompleteness Theorem.

The third question, ``Should we accept it as an axiom?'', moves us into
the realm of philosophy.  Today there are three major schools of thought
concerning the use of AC:

\begin{enumerate}
  \item Accept it as an axiom and use it without hesitation.
  \item Accept it as an axiom but use it only when you cannot find a
  proof without it.
  \item AC is unacceptable.
\end{enumerate}

Most mathematicians today belong to school A.  Mathematicians who are in
school B are usually there because of a belief in Occam's Razor (use as
few assumptions as possible when explaining something) or an interest in
metamathematics.  There are a growing number of people moving to school
C, especially computer scientists who work on automated reasoning using
constructive type theories.

Underlying the schools of thought about the use of AC are views about
truth and the nature of mathematical objects.  Three major views are
platonism, constructivism, and formalism.

\subsubsection{Platonism}

A platonist believes that mathematical objects exist independent of the
human mind, and a mathematical statement, such as AC, is objectively
either true or false.  A platonist accepts AC only if it is objectively
true, and probably falls into school A or C depending on her belief.  If
she isn't sure about AC's truth then she may be in school B so that once
she finds out the truth about AC she will know which theorems are true.

\subsubsection{Constructivism}

A constructivist believes that the only acceptable mathematical objects
are ones that can be constructed by the human mind, and the only
acceptable proofs are constructive proofs.  Since AC gives no method for
constructing a choice set constructivists belong to school C.

\subsubsection{Formalism}

A formalist believes that mathematics is strictly symbol manipulation
and any consistent theory is reasonable to study.  For a formalist the
notion of truth is confined to the context of mathematical models, e.g.,
a formalist would say ``The parallel postulate is false in Riemannian
geometry.'' but she wouldn't say ``The parallel postulate is false.''  A
formalist will probably not allign herself with any school.  She will
comfortably switch between A, B, and C depending on her current
interests.


So: Should you accept the Axiom of Choice?  Here are some arguments for
and against it.

\subsubsection{Against}

\begin{itemize}
  \item It's not as simple, aesthetically pleasing, and intuitive as the
  other axioms.
  \item It is equivalent to many statements which are not intuitive such
  as ``Every set can be well ordered.''  How, for example, would you
  well order the reals?
  \item With it you can derive non-intuitive results, such as the
  existence of a discontinuous additive function, the existence of a
  non-measurable set of reals, and the Banach-Tarski Paradox (see the
  next section of the sci.math FAQ).
  \item It is nonconstructive---it conjures up a set without providing
  any sort of procedure for its construction.
\end{itemize}

\subsubsection{For}

The acceptance of AC is based on the belief that our intuition about
finite sets can be extended to infinite sets.  The main argument for
accepting it is that it is useful.  Many important, intuitively
plausible theorems are equivalent to it or depend on it.  For example
these statements are equivalent to AC:
\begin{itemize}
  \item Every vector space has a basis.
  \item Trichotomy of Cardinals: For any cardinals $k$ and $l$, either
  $k<l$ or $k=l$ or $k>l$.
  \item Tychonoff's Theorem: The product of compact spaces is compact in
  the product topology.
  \item Zorn's Lemma: Every nonempty partially ordered set P in which
  each chain has an upper bound in P has a maximal element.
\end{itemize}

And these statements depend on AC (i.e., they cannot be proved in ZF
without AC):

\begin{itemize}
  \item The union of countably many countable sets is countable.
  \item Every infinite set has a denumerable subset.
  \item The Loewenheim-Skolem Theorem: Any first-order theory which has
  a model has a denumerable model.
  \item The Baire Category Theorem: The reals are not the union of
  countably many nowhere dense sets (i.e., the reals are not meager).
  \item The Ultrafilter Theorem: Every Boolean algebra has an
  ultrafilter on it.
\end{itemize}

\subsubsection{Alternatives to AC}

\begin{itemize}
  \item Accept only a weak form of AC such as the Denumerable Axiom of
  Choice (every denumerable set has a choice function) or the Axiom of
  Dependent Choice.
  \item Accept an axiom that implies AC such as the Axiom of
  Constructibility ($V=L$) or the Generalized Continuum Hypothesis
  (GCH).
  \item Adopt AC as a logical axiom (Hilbert suggested this with his
  epsilon axiom).  If set theory is done in such a logical formal system
  the Axiom of Choice will be a theorem.
  \item Accept a contradictory axiom such as the Axiom of Determinacy.
  \item Use a completely different framework for mathematics such as
  Category Theory.  Note that within the framework of Category Theory
  Tychonoff's Theorem can be proved without AC (Johnstone, 1981).
\end{itemize}

\subsubsection{Test Yourself: When is AC necessary?}

If you are working in Zermelo-Fraenkel set theory without the Axiom of
Choice, can you choose an element from...
\begin{enumerate}
  \item a finite set?
  \item an infinite set?
  \item each member of an infinite set of singletons (i.e., one-element
  sets)?
  \item each member of an infinite set of pairs of shoes?
  \item each member of inifinite set of pairs of socks?
  \item each member of a finite set of sets if each of the members is
  infinite?
  \item each member of an infinite set of sets if each of the members is
  infinite?
  \item each member of a denumerable set of sets if each of the members
  is infinite?
  \item each member of an infinite set of sets of rationals?
  \item each member of a denumerable set of sets if each of the members
  is denumberable?
  \item each member of an infinite set of sets if each of the members is
  finite?
  \item each member of an infinite set of finite sets of reals?
  \item each member of an infinite set of sets of reals?
  \item each member of an infinite set of two-element sets whose members
  are sets of reals?
\end{enumerate}
The answers to these questions with explanations are accessible through
\url{http://www.jazzie.com/ii/math/index.html}.

\Ref

Benacerraf, Paul and Putnam, Hilary.  ``Philosophy of Mathematics:
Selected Readings'', 2nd edition. Cambridge University Press, 1983.

Dauben, Joseph Warren.  ``Georg Cantor: His Mathematics and Philosophy
of the Infinite.''  Princeton University Press, 1979.

A. Fraenkel, Y.  Bar-Hillel, and A. Levy with van Dalen, Dirk.
``Foundations of Set Theory,'' Second Revised Edition.  North-Holland,
1973.

Johnstone, Peter T.  ``Tychonoff's Theorem without the Axiom of
Choice.''  Fundamenta Mathematica 113: 21-35, 1981.

Leisenring, Albert C.  ``Mathematical Logic and Hilbert's
Epsilon-Symbol.''  Gordon and Breach, 1969.

Maddy, ``Believing the Axioms, I'', J. Symb. Logic, v. 53, no. 2, June
1988, pp. 490-500, and ``Believing the Axioms II'' in v.53, no. 3.

Moore, Gregory H.  ``Zermelo's Axiom of Choice: Its Origins,
Development, and Influence.''  Springer-Verlag, 1982.

Rubin, Herman and Rubin, Jean E.  ``Equivalents of the Axiom of Choice
II.''  North-Holland, 1985.

This section of the FAQ is Copyright (c) 1994 Nancy McGough.  Send
comments and or corrections relating to this part to
\url{nancym@ii.com}.  The most up to date version of this section of the
sci.math FAQ is accesible through
\url{http://www.jazzie.com/ii/math/index.html}.
