The following books have been recommended by several readers.
The number of recommendations is in brackets.

\begin{itemize}
  \item Algebra

  Lang, Serge. Algebra. 2nd ed. Addison-Wesley Pub. Co., 1984. [2]

  Halmos, Linear Algebra. [1]

  Birkhoff, McLane, Algebra

  van der Waerden. Algebra

  Atiyah \& MacDonald. Introduction to Commutative Algebra

  \item Complex Analysis

  Ahlfors, Lars Valerian. Complex analysis: an introduction to the
  theory of analytic functions of one complex variable. 3rd ed. New
  York; Toronto: McGraw-Hill, c1979. [2]

  Conway, John B. Functions of one complex variable [by] John
  B. Conway. [New York] Springer-Verlag New York, 1973. [1]

  Priestley, Introduction to complex analysis

  \item Real \& Complex Analysis

  Titchmarsh. Theory of Functions

  Boas. Primer of Real Functions

  Polya \& Szego. Problems \& Theorems in Analysis

  Rudin, Walter. Principles of mathematical analysis. 3d ed. New York:
  McGraw-Hill, 1976. [2]

  Rudin, Walter. ``Functional Analysis''

  Royden, H. L. Real analysis. 3rd ed. New York, Macmillan; London:
  Collier Macmillan, 1988. [1]

  Hewitt, Edwin, 1920. ``Real and abstract analysis: a modern treatment
  of the theory of functions of a real variable.'' New York:
  Springer-Verlag, 1969. [2]

  Dieudonn\'e. Foundations of Analysis

  Courant \& Hilbert. Mathematical Methods of Physics.

  \item Geometry

  David Hilbert. {\em Foundations of Geometry} 2nd English edition,
  tr. by Leo Unger, publ. by Open Court, 1971.

  Neumann, Stoy \& Thompson. Groups and Geometry [1]

  \item Number Theory

  Hardy, Littlewood.

  Samuel, ``Algebraic Theory of Numbers''

  Hardy \& Wright

  \item History of Mathematics

  Morris Kline, Mathematical Thought from Ancient to Modern Times

  \item Topology

  Guillemin, Victor and Alan Pollack: Differential Topology.

  Spivak, Michael: A Comprehensive Introduction to Differential
  Geometry, Vol. I

  Morgan, Frank: Riemannian Geometry: A Beginner's Guide

  Milnor, ``Topology from the Differentiable Viewpoint''

  R. Engelking. General Topology.

  Kuratowski. Topology.

  Copson. Metric Spaces.

  Greenberg, Martin and (?) Harper: Algebraic Topology: An Introduction.

  Kelly, General topology

  \item Calculus

  Hardy, Course of Pure Mathematics.[2]

  Landau. Differential \& Integral Calculus.

  Courant \& John. Introduction to Calculus \& Analysis, vol.1.

  Spivak. Calculus on Manifolds.

  \item Probability

  Feller, Introduction to probability theory

  \item Statistics

  Silvey, Statistical inference

  \item Measure Theory

  Weir, Integration and measure

  \item General

  Courant \& Robins [2] What is Mathematics.  Oxford University Press.
  1969
\end{itemize}
%%% Local Variables:
%%% mode: latex
%%% TeX-master: "math-faq"
%%% End:
