


A basic reference is Godel's ``What is Cantor's Continuum Problem?", from
1947 with a 1963 supplement, reprinted in Benacerraf and Putnam's collection
Philosophy of Mathematics.  This outlines Godel's generally anti-CH views,
giving some ``implausible" consequences of CH.

  "I believe that adding up all that has been said one has good reason
   to suspect that the role of the continuum problem in set theory will
   be to lead to the discovery of new axioms which will make it possible
   to disprove Cantor's conjecture."

At one stage he believed he had a proof that $C = \aleph_2$ from some new
axioms, but this turned out to be fallacious.  (See Ellentuck, ``Godel's
Square Axioms for the Continuum", Mathematische Annalen 1975.)

Maddy's  ``Believing the Axioms",
Journal of Symbolic Logic 1988 (in 2 parts) is an extremely
interesting paper and a lot of fun to read.  A bonus is that it gives
a non-set-theorist who knows the basics a good feeling for a lot of
issues in contemporary set theory.

Most of the first part is devoted to ``plausible arguments" for or against
CH: how it stands relative to both other possible axioms and to various
set-theoretic ``rules of thumb".  One gets the feeling that the weight of
the arguments is against CH, although Maddy says that many ``younger members"
of the set-theoretic community are becoming more sympathetic to CH than
their elders.  There's far too much here for me to be able to go into it
in much detail.

Some highlights from Maddy's discussion, also incorporating a few things that
other people sent me:

\begin{enumerate}
\item Cantor's reasons for believing CH aren't all that persuasive today.
\item Godel's proof of the consistency of CH shows that CH follows from
    ZFC plus the Axiom of Constructibility ($V=L$, roughly that the set-theoretic
    universe = the constructible universe).  However, most set-theorists
    seem to find Constructiblity implausible and much too restrictive.  It's
    an example of a ``minimizing" principle, which tends to cut down on
    the number of sets admitted to one's universe.  Apparently ``maximizing"
    principles meet with much more sympathy from set theorists.  Such
    principles are more compatible with $\neg$CH than with CH.
\item If GCH is true, this implies that $\aleph_0$ has certain unique properties:
    e.g. that it's that cardinal before which GCH is false and after which
    it is true.  Some would like to believe that the set-theoretic universe
    is more ``uniform" (homogeneous) than that, without this kind of
    singular occurrence.  Such a ``uniformity" principle tends to imply $\neg$GCH.
\item Most of those who disbelieve CH think that the continuum is likely to
    have very large cardinality, rather than $\aleph_2$ (as Godel seems to
    have suggested).  Even Cohen, a professed formalist, argues that the
    power set operation is a strong operation that should yield sets
    much larger than those reached quickly by stepping forward through the
    ordinals:
\begin{quote}
      "This point of view regards C as an incredibly rich set given to us
       by a bold new axiom, which can never be approached by any piecemeal
       process of construction."
\end{quote}
\item There are also a few arguments in favour of CH, e.g. there's an argument
    that $\neg$CH is restrictive (in the sense of (2) above).  Also, CH is
    much easier to force (Cohen's method) than $\neg$CH.  And CH is much more
    likely to settle various outstanding results than is $\neg$CH, which tends
    to be neutral on these results.
\item Most large cardinal axioms (asserting the existence of cardinals with
    various properties of hugeness: these are usually derived either from
    considering the hugeness of $\aleph_0$ compared to the finite cardinals
    and applying uniformity, or from considering the hugeness of V (the
    set-theoretic universe) relative to all sets and applying ``reflection")
    don't seem to settle CH one way or the other.
\item Various other axioms have some bearing.  Axioms of determinacy restrict
    the class of sets of reals that might be counterexamples to CH.  Various
    forcing axioms (e.g. Martin's axiom), which are ``maximality" principles
    (in the sense of (2) above), imply $\neg$CH.  The strongest (Martin's maximum)
    implies that $C = \aleph_2$.  Of course the ``truth" or otherwise of all
    these axioms is controversial.
\item Freiling's principle about ``throwing darts at the real line" is a
    seemingly very plausible principle, not involving large cardinals at
    all, from which $\neg$CH immediately follows.  Freiling's paper (JSL 1986)
    is a good read.  More on this at the end of this message.
\end{enumerate}

Of course we have conspicuously avoided saying anything about whether it's
even reasonable to suppose that CH has a determinate truth-value.
Formalists will argue that we may choose to make it come out whichever
way we want, depending on the system we work in.  On the other hand,
the mere fact of its independence from ZFC shouldn't immediately lead
us to this conclusion -- this would be assigning ZFC a privileged status
which it hasn't necessarily earned.  Indeed, Maddy points out that
various axioms within ZFC (notably the Axiom of Choice, and also
Replacement) were adopted for extrinsic reasons (e.g. ``usefulness")
as well as for ``intrinsic" reasons (e.g. ``intuitiveness").  Further
axioms, from which CH might be settled, might well be adopted for such
reasons.

One set-theorist correspondent said that set-theorists themselves are very
loathe to talk about ``truth" or ``falsity" of such claims.  (They're
prepared to concede that $2+2=4$ is true, but as soon as you move beyond
the integers trouble starts.  e.g. most were wary even of suggesting that
the Riemann Hypothesis necessarily has a determinate truth-value.)  On
the other hand, Maddy's contemporaries discussed in her paper seemed quite
happy to speculate about the ``truth" or ``falsity" of CH.

The integers are not only a bedrock, but also
any finite number of power sets seem to be quite natural
Intuitively are also natural which would point towards the fact
that  CH may be determinate
one way or the other.  As one correspondent suggested, the question of the
determinateness of CH is perhaps the single best way to separate the
Platonists from the formalists.


 And is it true or false?  Well, CH is somewhat intuitively
plausible.  But after reading all this, it does seem that the weight of
evidence tend to point the other way.

The following is from Bill Allen on Freiling's Axiom of Symmetry.
This is a good one to run your intuitions by.
\begin{quote}
Let $A$ be the set of functions mapping Real Numbers into countable sets
of Real Numbers.  Given a function $f$ in $A$, and some arbitrary real
numbers $x$ and $y$, we see that $x$ is in $f(y)$ with probability 0, i.e. $x$
 is
not in $f(y)$ with probability 1.  Similarly, $y$ is not in $f(x)$ with
probability 1.  Let AX be the axiom which states

     ``for every $f$ in $A$, there exist $x$ and $y$ such that $x$ is not in 
$f(y)$
      and $y$ is not in $f(x)$"

The intuitive justification for AX is that we can find the $x$ and $y$ by
choosing them at random.

In ZFC, AX = not CH.
proof:
If CH holds, then well-order $R$ as $r_0, r_1, .... , r_x, ...$
with $x < \aleph_1$.  Define $f(r_x)$ as $\{r_y : y \leq x\}$.  Then $f$ is a
function which witnesses the falsity of AX.

If CH fails, then let $f$ be some member of $A$.  Let $Y$ be a subset of $R$ of
cardinality $\aleph_1$.  Then $Y$ is a proper subset.  Let $X$ be the union of
all the sets $f(y)$ with $y$ in $Y$, together with $Y$.  Then, as $X$ is an
$\aleph_1$ union of countable sets, together with a single $\aleph_1$ size set
$Y$, the cardinality of $X$ is also $\aleph_1$, so $X$ is not all of $R$. 
 Let a be
in $R \ X$, so that a is not in $f(y)$ for any $y$ in $Y$.  Since $f(a)$ is
countable, there has to be some $b$ in $Y$ such that $b$ is not in $f(a)$. 
 Thus
we have shown that there must exist $a$ and $b$ such that $a$ is not in $f(b)$
and $b$ is not in $f(a)$.  So AX holds. 
\end{quote}
Freiling's proof, does not invoke large cardinals or
intense infinitary combinatorics to make the point that CH implies
counter-intuitive propositions.  Freiling has also pointed out that the
natural extension of AX is AXL (notation mine), where AXL is AX with
the notion of countable replaced by Lebesgue Measure zero.  Freiling has
established some interesting Fubini-type theorems using AXL.

See ``Axioms of Symmetry: Throwing Darts at the Real Line", by Freiling,
Journal of Symbolic Logic, 51, pages 190-200.  An extension of this work
appears in "Some properties of large filters", by Freiling and Payne, in
the JSL, LIII, pages 1027-1035.

The section above was excerpted from a posting from David Chalmers,
of Indiana University.


See also

   Nancy McGough's *Continuum Hypothesis article* or its *mirror*.


 http://www.jazzie.com/ii/math/ch/


 http://www.best.com/~ii/math/ch/

