\section{Erd\H os Number}

Form an undirected graph where the vertices are academics, and an edge
connects academic $X$ to academic $Y$ if $X$ has written a paper with
$Y$. The Erd\H os number of $X$ is the length of the shortest path in
this graph connecting $X$ with Erd\H os.

Erd\H os has Erd\H os number $0$.  Co-authors of Erd\H os have Erd\H os
number $1$.  Einstein has Erd\H os number $2$, since he wrote a paper
with Ernst Straus, and Straus wrote many papers with Erd\H os.


The Extended Erd\H os Number applies to co-authors of Erd\H os.  For
People who have authored more than one paper with Erd\H os, their Erd\H
os number is defined to be $1/\#$ papers-co-authored.
%  Ron Graham has the smallest, non-zero, Erd\H os number.

Why people care about it?

Nobody seems to have a reasonable answer...

\subsection{Who is Paul Erd\H os?}

Paul Erd\H os was an Hungarian mathematician. He obtained his Ph.D.
from the University of Manchester and spent most of his efforts tackling
``small'' problems and conjectures related to graph theory,
combinatorics, geometry and number theory.

He was one of the most prolific publishers of papers; and was also and
indefatigable traveller.

Paul Erd\H os died on September 20, 1996.

At this time the number of people with Erd\H os number $2$ or less is
estimated to be over 4750, according to Professor Jerrold W. Grossman
archives. These archives can be consulted via anonymous ftp at
\url{ftp://vela.acs.oakland.edu/pub/math/erdos} or on the Web at
\url{http://www.acs.oakland.edu/~grossman/erdoshp.html}.  At this time
it contains a list of all co-authors of Erd\H os and their co-authors.

On this topic, he writes

\begin{quote}
  Let $E_1$ be the subgraph of the collaboration graph induced by people
  with Erd\H os number~$1$. We found that $E_1$ has 451 vertices and
  1145 edges.  Furthermore, these collaborators tended to collaborate a
  lot, especially among themselves. They have an average of 19 other
  collaborators (standard deviation~$21$), and only seven of them
  collaborated with no one except Erd\H os. Four of them have over 100
  co-authors. If we restrict our attention just to~$E_1$, we still find
  a lot of joint work. Only 41 of these 451 people have collaborated
  with no other persons with Erd\H os number~$1$ (i.e., there are 41
  isolated vertices in~$E_1$), and $E_1$ has four components with two
  vertices each. The remaining 402 vertices in $E_1$ induce a connected
  subgraph. The average vertex degree in $E_1$ is~$5$, with a standard
  deviation of~$6$; and there are four vertices with degrees of 30 or
  higher. The largest clique in $E_1$ has seven vertices, but it should
  be noted that six of these people and Erd\H os have a joint
  seven-author paper. In addition, there are seven maximal 6-cliques,
  and 61 maximal 5-cliques. In all, 29 vertices in $E_1$ are involved in
  cliques of order 5 or larger. Finally, we computed that the diameter
  of $E_1$ is 11 and its radius is~6.

  Three quarters of the people with Erd\H os number~$2$ have only one
  co-author with Erd\H os number~$1$ (i.e., each such person has a
  unique path of length $2$ to~$p$). However, their mean number of Erd\H
  os number~$1$ co-authors is $1.5$, with a standard deviation of~$1.1$,
  and the count ranges as high as~$13$.

  Folklore has it that most active researchers have a finite, and fairly
  small, Erd\H os number. For supporting evidence, we verified that all
  the Fields and Nevanlinna prize winners during the past three cycles
  ($1986--1994$) are indeed in the Erd\H os component, with Erd\H os
  number at most~$9$. Since this group includes people working in
  theoretical physics, one can conjecture that most physicists are also
  in the Erd\H os component, as are, therefore, most scientists in
  general. The large number of applications of graph theory to the
  social sciences might also lead one to suspect that many researchers
  in other academic areas are included as well. We close with two open
  questions about~$C$, restricted to mathematicians, that such musings
  suggest, with no hope that either will ever be answered
  satisfactorily: What is the diameter of the Erd\H os component, and
  what is the order of the second largest component?
\end{quote}

\Ref

\article{Caspar Goffman.}{And what is your Erd\H os number?}  {American
  Mathematical Monthly,}{v. 76 (1969), p. 791.}

\article{Tom Odda (alias for Ronald Graham)} {On Properties of a
  Well-Known Graph, or, What is Your Ramsey Number?}  { Topics in Graph
  Theory}{ (New York, 1977), pp.  [166-172].}
