\section{Fields Medal}

\subsection{Historical Introduction}

This is the original letter by Fields creating the endowment for the
medals that bear his name.  It is thought to have been written during
the few months before his death.  Notice that no mention is made about
the age of the recipients (currently there is a 40 year-old limit), and
that the medal should not be attached to any person, private or public,
meaning that it shouldn't bear anybody's name.

Colloquially, people refer to the Fields Medal as ``the Nobel Prize for
Mathematics''. This is not quite correct, as the age restriction makes
it impossible to award the prize to distinguished mathematicians who are
very deserving, such as Andrew Wiles.

\begin{quote}
  It is proposed to found two gold medals to be awarded at successive
  International Mathematical Congress for outstanding achievements in
  mathematics. Because of the multiplicity of the branches of
  mathematics and taking into account the fact that the interval between
  such congresses is four years it is felt that at least two medals
  should be available. The awards would be open to the whole world and
  would be made by an International Committee.

  The fund for the founding of the medals is constituted by balance left
  over after financing the Toronto congress held in 1924. This must be
  held in trust by the Government or by some body authorized by
  government to hold and invest such funds. It would seem that a
  dignified method for handling the matter and one which in this
  changing world should most nearly secure permanency would be for the
  Canadian Government to take over the fund and appoint as his custodian
  say the Prime Minister of the Dominion or the Prime Minister in
  association with the Minister of Finance. The medals would be struck
  at the Mint in Ottawa and the duty of the custodian would be simply to
  hand over the medals at the proper time to the accredited
  International Committee.

  As things are at present a practical course of procedure would seem to
  be for the Executive Committee of a Congress to appoint a small
  international committee authorized to add to its number and call into
  consultation other mathematicians as it might deem expedient. The
  Committee would be expected to decide on the ones to whom the awards
  should be made thirty months in advance of the following Congress. Its
  decisions would be communicated to the President and Secretary of the
  Organizing Committee of the Congress, this Committee having the duty
  of communicating to the Prime Minister of Canada the names of the
  recipients in order that the medal might be prepared in time and
  forwarded to the president of the Organizing Committee. Immediately on
  the appointment of the Executive Committee of the Congress the medals
  would be handed over to its President. The presentation of the medals
  would constitute a special feature at some general meeting of the
  Congress.

  In the above arrangements the role of the Organizing Committee might
  be taken over by the Executive of the International Mathematical Union
  at some time in the future when that organization has been generally
  accepted.

  In coming to its decision the hands of the IC should be left as free
  as possible. It would be understood, however, that in making the
  awards while it was in recognition of work already done it was at the
  same time intended to be an encouragement for further achievement on
  the part of the recipients and a stimulus to renewed effort on the
  part of others.

  In commenting on the work of the medalists it might be well to be
  conservative in one's statements to avoid envidious comparisons
  explicit or implied. The Committee might ease matters by saying they
  have decided to make the awards along certain lines not alone because
  of the outstanding character of the achievement but also with a view
  to encouraging further development along these lines. In this
  connection the Committee might say that they had elected to select
  subjects in Analysis, in Geometry, in the Theory of Groups, in the
  Theory of Numbers etc. as the case might be. When the Committee had
  come to an agreement in this sense the claims for recognition of work
  done along the special lines in question could be considered in detail
  by two smaller groups or subcommittees with specialized qualifications
  who would have authority to take into consultation or add to the
  subcommittees other mathematicians of specialized knowledge.

  With regard to the medals themselves, I might say that they should
  each contain at least 200 dollars worth of gold and be of a fair size,
  probably 7.5 centimeters in diameter. Because of the international
  character the language to be employed it would seem should be Latin or
  Greek? The design has still to be definitely determined. It will have
  to be decided on by artists in consultation with mathematicians. The
  suggestions made in the preceding are tentative and open to
  consideration on the part of mathematicians.

  It is not contemplated to make an award until 1936 at the Congress
  following that at Zurich during which an international Medal Committee
  should be named.

  The above programme means a new departure in the matter of
  international scientific cooperation and is likely to be the precursor
  of moves along like lines in other sciences than mathematics.

  One would hear again emphasized the fact that the medals should be of
  a character as purely international and impersonal as possible. There
  should not be attached to them in any way the name of any country,
  institution or person.

  Perhaps provision could be made as soon as possible after the
  appointment of the Executive of the Zurich Congress for the
  consideration by it of the subject of the medals, and the appointment
  without undue delay of a Committee and the awards of the medals to be
  made in connection with the Congress of 1936.

  Suggestions with regard to the design of the medals will be welcome.


                                 (signed)   J.C. Fields\\
                                            Research Professor of Mathematics\\
                                            University of Toronto
\end{quote}


More information may also be found at
%
%From Joel Chan
%Date: Sat, 25 Nov 1995 22:54:20 -0500
%on your next update could you change the URL
%
\url{http://www.math.toronto.edu/fields.html}

\vfill\pagebreak

\subsection{Table of Awardees}

\begin{tabular}{|r|l|l|l|r|} \hline

  {Year} & {Name} & {Birthplace} & {Country}& {Age}\\
  \hline

1936&Ahlfors, Lars       &Helsinki       &Finland   &29\\
1936&Douglas, Jesse      &New York, NY   &USA       &39\\
1950&Schwartz, Laurent   &Paris          &France    &35\\
1950&Selberg, Atle       &Langesund      &Norway    &33\\
1954&Kodaira, Kunihiko   &Tokyo          &Japan     &39\\ 
1954&Serre, Jean-Pierre  &Bages          &France    &27\\
1958&Roth, Klaus         &Breslau        &Germany   &32\\
1958&Thom, Rene          &Montbeliard    &France    &35\\
1962&Hormander, Lars     &Mjallby        &Sweden    &31\\
1962&Milnor, John        &Orange, NJ     &USA       &31\\
1966&Atiyah, Michael     &London         &UK        &37\\
1966&Cohen, Paul         &Long Branch NJ &USA       &32\\
1966&Grothendieck, Alex. &Berlin         &Germany   &38\\
1966&Smale, Stephen      &Flint, MI      &USA       &36\\
1970&Baker, Alan         &London         &UK        &31\\
1970&Hironaka, Heisuke   &Yamaguchi-ken  &Japan     &39\\
1970&Novikov, Serge      &Gorki          &USSR      &32\\
1970&Thompson, John      &Ottawa, KA     &USA       &37\\
1974&Bombieri, Enrico    &Milan          &Italy     &33\\
1974&Mumford, David      &Worth, Sussex  &UK        &37\\
1978&Deligne, Pierre     &Brussels       &Belgium   &33\\
1978&Fefferman, Charles  &Washington DC  &USA       &29\\
1978&Margulis, Gregori   &Moscow         &USSR      &32\\
1978&Quillen, Daniel     &Orange, NJ     &USA       &38\\
1982&Connes, Alain       &Draguignan     &France    &35\\
1982&Thurston, William   &Washington DC  &USA       &35\\
1982&Yau, Shing-Tung     &Kwuntung       &China     &33\\
1986&Donaldson, Simon    &Cambridge      &UK        &27\\
1986&Faltings, Gerd      &1954           &Germany   &32\\
1986&Freedman, Michael   &Los Angeles    &USA       &35\\
\hline
\end{tabular}

\begin{tabular}{|r|l|l|l|r|} \hline

{Year} & {Name} & {Birthplace} & {Country}& {Age}\\
\hline

1990&Drinfeld, Vladimir  &Kharkov       &USSR      &36\\
1990&Jones, Vaughan      &Gisborne      &N Zealand &38\\
1990&Mori, Shigefumi     &Nagoya        &Japan     &39\\
1990&Witten, Edward      &Baltimore     &USA       &38\\
1994& Pierre-Louis Lions    & ????      &France    &38\\
1994&Jean-Chrisophe Yoccoz  &   ????    &France    &37\\
%From: Elly Gustafsson <gustafss@IAS.EDU>
%Date: Tue, 19 Mar 1996 16:37:31 -0500
%Birthplace:   Oostende
%Country:      Belgium
%Institution:  Institute for Advanced Study   only
1994&Jean Bourgain       &Oostende       &Belge    &40\\
%Date: Tue, 23 Apr 1996 12:13:09 -0400
%From: Dmitriy Rumynin <rumynin@math.umass.edu>
%The place of birth of Efim Zelmanov was Novosibirsk.
1994&Efim Zelmanov       &Novosibirsk    &Russia   &39\\
\hline
\end{tabular}

\bigskip
\bigskip

\begin{tabular}{|r|l|l|l|} \hline

{Year} & {Name} & {Institution} &{Country}\\
\hline

1936&Ahlfors, Lars       &Harvard University         &USA \\
1936&Douglas, Jesse      &MIT                        &USA \\
1950&Schwartz, Laurent   &Universite de Nancy        &France \\
1950&Selberg, Atle       &Institute for Advanced Study, Princeton&USA \\
1954&Kodaira, Kunihiko   &Princeton University       &USA \\
1954&Serre, Jean-Pierre  &College de France          &France \\
1958&Roth, Klaus         &University of London       &UK \\
1958&Thom, Rene          &University of Strasbourg   &France \\
1962&Hormander, Lars     &University of Stockholm    &Sweden \\
1962&Milnor, John        &Princeton University       &USA \\
1954&Serre, Jean-Pierre  &College de France          &France \\
1958&Roth, Klaus         &University of London       &UK \\
1958&Thom, Rene          &University of Strasbourg   &France \\
1962&Hormander, Lars     &University of Stockholm    &Sweden \\
1962&Milnor, John        &Princeton University       &USA \\
1966&Atiyah, Michael     &Oxford University          &UK \\
1966&Cohen, Paul         &Stanford University        &USA \\
1966&Grothendieck, Alex  &University of Paris        &France \\
1966&Smale, Stephen      &University of California at Berkeley&USA \\
1970&Baker, Alan         &Cambridge University       &UK \\
1970&Hironaka, Heisuke   &Harvard University         &USA \\
1970&Novikov, Serge      &Moscow University          &USSR \\
1970&Thompson, John      &University of Chicago      &USA \\
1974&Bombieri, Enrico    &Univeristy of Pisa         &Italy \\
1974&Mumford, David      &Harvard University         &USA \\
1978&Deligne, Pierre     &IHES                       &France \\
1978&Fefferman, Charles  &Princeton University       &USA \\
1978&Margulis, Gregori   &InstPrblmInfTrans          &USSR \\
1978&Quillen, Daniel     &MIT                        &USA \\
1982&Connes, Alain       &IHES                       &France \\
1982&Thurston, William   &Princeton University       &USA \\
1982&Yau, Shing-Tung     &Institute for Advanced Study, Princeton&USA \\
1986&Donaldson, Simon    &Oxford University          &UK \\
1986&Faltings, Gerd      &Princeton University       &USA \\
1986&Freedman, Michael   &University of California at San Diego&USA \\
\hline
\end{tabular}

\begin{tabular}{|r|l|l|l|} \hline

{Year} & {Name} & {Institution} &{Country}\\
\hline

1990&Drinfeld, Vladimir  &Phys.Inst.Kharkov          &USSR \\
1990&Jones, Vaughan      &University of California at Berkeley&USA \\
1990&Mori, Shigefumi     &University of Kyoto?       &Japan \\
1990&Witten, Edward      &Princeton/Institute for Advanced Study&USA \\
1994&Lions, Pierre-Louis &Universite de Paris-Dauphine&France\\
1994&Yoccoz, Jean-Chrisophe&Universite de Paris-Sud  &France\\
1994&Bourgain, Jean      &Institute for Advanced Study&USA\\
1994&Zelmanov, Efim      &University of Wisconsin    &USA\\
1998&Borcherd, Richard   &Cambridge University       &Great Britain \\
1998&Gowers, William Timothy&Cambridge University    &Great Britain\\
1998&Kontsevich, Maxim   &IHES Bures-sur-Yvette      &Russia\\
1998&McCullen, Curtis    &Harvard University         &USA\\
\hline
\end{tabular}

\bigskip
\bigskip


\Ref

\book{International Mathematical Congresses, An Illustrated History
  1893-1986.}  {Donald J.Alberts, G. L. Alexanderson and Constance
  Reid.}  {Revised Edition, Including 1986, Springer Verlag, 1987.}

\article{Tropp, Henry S.}  {The origins and history of the Fields
  Medal.}  {Historia Mathematica,} {3(1976), 167-181.}
%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "math-faq"
%%% End: 
