\section{Fermat's Last Theorem}

\subsection{History of Fermat's Last Theorem}

Pierre de Fermat (1601-1665) was a lawyer and amateur mathematician. In
about 1637, he annotated his copy (now lost) of Bachet's translation of
Diophantus' Arithmetika with the following statement:

\begin{quote}
  {\it Cubum autem in duos cubos, aut quadratoquadratum in duos
    quadratoquadratos, et generaliter nullam in infinitum ultra
    quadratum potestatem in duos ejusdem nominis fas est dividere: cujus
    rei demonstrationem mirabilem sane detexi. Hanc marginis exiguitas
    non caperet.}
\end{quote}

In English, and using modern terminology, the paragraph above reads as:
\begin{quote}
  There are no positive integers such that $x^n + y^n = z^n$ for $n>2$.
  I've found a remarkable proof of this fact, but there is not enough
  space in the margin [of the book] to write it.
\end{quote}

Fermat never published a proof of this statement. It became to be known
as Fermat's Last Theorem (FLT) not because it was his last piece of
work, but because it is the last remaining statement in the post-humous
list of Fermat's works that needed to be proven or independently
verified. All others have either been shown to be true or disproven long
ago.


\subsection{What is the current status of FLT?}

\bigskip

\begin{teo}[Fermat's Last Theorem]
  There are no positive integers $x$, $y$, $z$, and $n > 2$ such that
  $x^n + y^n = z^n$.
\end{teo}

Andrew Wiles, a researcher at Princeton, claims to have found a proof.
The proof was presented in Cambridge, UK during a three day seminar to
an audience which included some of the leading experts in the field.
The proof was found to be wanting.  In summer 1994, Prof. Wiles
acknowledged that a gap existed. On October 25th, 1994, Prof. Andrew
Wiles released two preprints, {\em Modular elliptic curves and Fermat's
  Last Theorem}, by Andrew Wiles, and {\em Ring theoretic properties of
  certain Hecke algebras}, by Richard Taylor and Andrew Wiles.  The
first one (long) announces a proof of, among other things, Fermat's Last
Theorem, relying on the second one (short) for one crucial step.

The argument described by Wiles in his Cambridge lectures had a serious
gap, namely the construction of an Euler system.  After trying
unsuccessfully to repair that construction, Wiles went back to a
different approach he had tried earlier but abandoned in favor of the
Euler system idea.  He was able to complete his proof, under the
hypothesis that certain Hecke algebras are local complete intersections.
This and the rest of the ideas described in Wiles' Cambridge lectures
are written up in the first manuscript.  Jointly, Taylor and Wiles
establish the necessary property of the Hecke algebras in the second
paper.

The new approach turns out to be significantly simpler and shorter than
the original one, because of the removal of the Euler system.  (In fact,
after seeing these manuscripts Faltings has apparently come up with a
further significant simplification of that part of the argument.)

% As indicated by Dana W. Albrecht dwa@corsair.com. Jul 21/1995.
% Correction added Aug 21/1995

The papers were published in the May 1995 issue of {\it Annals of
  Mathematics}.  For single copies of the issues send e-mail to
\url{jlorder@jhunix.hcf.jhu.edu} for further directions.

In summary:

Both manuscripts have been published. Thousands of people have a read
them.  About a hundred understand it very well. Faltings has simplified
the argument already. Diamond has generalized it. People can read
it. The immensely complicated geometry has mostly been replaced by
simpler algebra. The proof is now generally accepted. There was a gap in
this second proof as well, but it has been filled since October 1994.

%Apparently retired without replacement, as noted by:
%From: wdr@world.std.com (William D Ricker)
%Date: Tue, 28 Nov 1995 13:37:56 -0500
%
%You may also peruse the AMS site on Fermat's Last Theorem at:
%
%\begin{verbatim}
%gopher://e-math.ams.org/11/lists/fermat
%\end{verbatim}

\CC{
  \subsection{Wiles' line of attack}

  Here is an outline of the {\bf first, incorrect} proposed proof.  The
  bits about Euler system are

  \begin{itemize}

    \item {\bf From Ken Ribet:}

    Here is a brief summary of what Wiles said in his three lectures.

    The method of Wiles borrows results and techniques from lots and
    lots of people.  To mention a few: Mazur, Hida, Flach, Kolyvagin,
    yours truly, Wiles himself (older papers by Wiles), Rubin...  The
    way he does it is roughly as follows.  Start with a mod $p$
    representation of the Galois group of $Q$ which is known to be
    modular.  You want to prove that all its lifts with a certain
    property are modular.  This means that the canonical map from
    Mazur's universal deformation ring to its maximal Hecke algebra
    quotient is an isomorphism.  To prove a map like this is an
    isomorphism, you can give some sufficient conditions based on
    commutative algebra.  Most notably, you have to bound the order of a
    cohomology group which looks like a Selmer group for $Sym^2$ of the
    representation attached to a modular form.  The techniques for doing
    this come from Flach; and then the proof went on to use Euler
    systems a la Kolyvagin, except in some new geometric guise. [This
    part turned out to be wrong and unnecessary].
    
    If you take an elliptic curve over $Q$, you can look at the
    representation of Gal on the 3-division points of the curve.  If
    you're lucky, this will be known to be modular, because of results
    of Jerry Tunnell (on base change).  Thus, if you're lucky, the
    problem I described above can be solved (there are most definitely
    some hypotheses to check), and then the curve is modular.
    Basically, being lucky means that the image of the representation of
    Galois on 3-division points is $GL(2,Z/3Z)$.
    
    Suppose that you are unlucky, i.e., that your curve $E$ has a
    rational subgroup of order 3.  Basically by inspection, you can
    prove that if it has a rational subgroup of order 5 as well, then it
    can't be semistable.  (You look at the four non-cuspidal rational
    points of $X_0(15)$.)  So you can assume that $E[5]$ is ``nice''.
    Then the idea is to find an $E^\prime$ with the same 5-division
    structure, for which $E^\prime[3]$ is modular.  (Then $E^\prime$ is
    modular, so $E^\prime[5] = E[5]$ is modular.)  You consider the
    modular curve $X$ which parameterizes elliptic curves whose
    5-division points look like $E[5]$.  This is a twist of $X(5)$.
    It's therefore of genus 0, and it has a rational point (namely,
    $E$), so it's a projective line.  Over that you look at the
    irreducible covering which corresponds to some desired 3-division
    structure.  You use Hilbert irreducibility and the Cebotarev density
    theorem (in some way that hasn't yet sunk in) to produce a
    non-cuspidal rational point of $X$ over which the covering remains
    irreducible.  You take $E^\prime$ to be the curve corresponding to
    this chosen rational point of $X$.
    
    
    \item {\bf From the previous version of the FAQ:}
    
    (b) conjectures arising from the study of elliptic curves and
    modular forms. -- The Taniyama-Weil-Shimura conjecture.
     
    There is a very important and well known conjecture known as the
    Taniyama-Weil-Shimura conjecture that concerns elliptic curves.
    This conjecture has been shown by the work of Frey, Serre, Ribet,
    et. al. to imply FLT uniformly, not just asymptotically as with the
    ABC conjecture.
    
    The conjecture basically states that all elliptic curves can be
    parameterized in terms of modular forms.

    There is new work on the arithmetic of elliptic curves. Sha, the
    Tate-Shafarevich group on elliptic curves of rank 0 or 1. By the way
    an interesting aspect of this work is that there is a close
    connection between Sha, and some of the classical work on FLT. For
    example, there is a classical proof that uses infinite descent to
    prove FLT for $n = 4$. It can be shown that there is an elliptic
    curve associated with FLT and that for $n=4$, Sha is trivial. It can
    also be shown that in the cases where Sha is non-trivial, that
    infinite-descent arguments do not work; that in some sense ``Sha
    blocks the descent''. Somewhat more technically, Sha is an
    obstruction to the local-global principle [e.g. the Hasse-Minkowski
    theorem].

    \item {\bf From Karl Rubin:}

    \begin{teo}
      If $E$ is a semistable elliptic curve defined over $Q$, then $E$
      is modular.
    \end{teo}

    It has been known for some time, by work of Frey and Ribet, that
    Fermat follows from this.  If $u^q + v^q + w^q = 0$, then Frey had
    the idea of looking at the (semistable) elliptic curve $y^2 =
    x(x-a^q)(x+b^q)$.  If this elliptic curve comes from a modular form,
    then the work of Ribet on Serre's conjecture shows that there would
    have to exist a modular form of weight 2 on $\Gamma_0(2)$.  But
    there are no such forms.
    
    To prove the Theorem, start with an elliptic curve $E$, a prime $p$
    and let
    \[\rho_p : Gal(\bar{Q}/Q) \rightarrow GL_2(Z/pZ)\]
    be the representation giving the action of Galois on the $p$-torsion
    $E[p]$.  We wish to show that a {\em certain} lift of this
    representation to $GL_2(Z_p)$ (namely, the $p$-adic representation
    on the Tate module $T_p(E)$) is attached to a modular form.  We will
    do this by using Mazur's theory of deformations, to show that {\em
      every} lifting which ``looks modular'' in a certain precise sense
    is attached to a modular form.
    
    Fix certain ``lifting data'', such as the allowed ramification,
    specified local behavior at $p$, etc. for the lift. This defines a
    lifting problem, and Mazur proves that there is a universal lift,
    i.e. a local ring $R$ and a representation into $GL_2(R)$ such that
    every lift of the appropriate type factors through this one.
    
    Now suppose that $\rho_p$ is modular, i.e. there is {\em some} lift
    of $\rho_p$ which is attached to a modular form.  Then there is also
    a hecke ring $T$, which is the maximal quotient of $R$ with the
    property that all {\em modular} lifts factor through $T$.  It is a
    conjecture of Mazur that $R = T$, and it would follow from this that
    {\em every} lift of $\rho_p$ which ``looks modular'' (in particular
    the one we are interested in) is attached to a modular form.
    
    Thus we need to know 2 things:

    \begin{enumerate}
      \item[(a)]  $\rho_p$ is modular
      \item[(b)]  $R = T$.
    \end{enumerate}
    
    It was proved by Tunnell that $\rho_3$ is modular for every elliptic
    curve.  This is because $PGL_2(Z/3Z) = S_4$.  So (a) will be
    satisfied if we take $p=3$.  This is crucial.
    
    Wiles uses (a) to prove (b) under some restrictions on $\rho_p$.
    Using (a) and some commutative algebra (using the fact that $T$ is
    Gorenstein, basically due to Mazur) Wiles reduces the statement $T =
    R$ to checking an inequality between the sizes of 2 groups.  One of
    these is related to the Selmer group of the symmetric square of the
    given modular lifting of $\rho_p$, and the other is related (by work
    of Hida) to an $L$-value.  The required inequality, which everyone
    presumes is an instance of the Bloch-Kato conjecture, is what Wiles
    needs to verify.
    
    [This is the part that turned out to be wrong in the first version].
    He does this using a Kolyvagin-type Euler system argument.  This is
    the most technically difficult part of the proof, and is responsible
    for most of the length of the manuscript.  He uses modular units to
    construct what he calls a {\it geometric Euler system} of cohomology
    classes.  The inspiration for his construction comes from work of
    Flach, who came up with what is essentially the bottom level of this
    Euler system.  But Wiles needed to go much farther than Flach did.
    In the end, {\em under certain hypotheses} on $\rho_p$ he gets a
    workable Euler system and proves the desired inequality.  Among
    other things, it is necessary that $\rho_p$ is irreducible.
    
    [The new proof replaces the argument above with one using
    commutative algebra and and some clever observations by De Shalit to
    fill in the gap.]

    Suppose now that $E$ is semistable.
    
    \noindent Case 1.  $\rho_3$ is irreducible.\\
    \noindent
    Take $p=3.$ By Tunnell's theorem (a) above is true.  Under these
    hypotheses the argument above works for $\rho_3$, so we conclude
    that $E$ is modular.
    
    \noindent Case 2.  $\rho_3$ is reducible.  Take $p=5$.  In this case
    $\rho_5$ must be irreducible, or else $E$ would correspond to a
    rational point on $X_0(15)$.  But $X_0(15)$ has only 4 noncuspidal
    rational points, and these correspond to non-semistable curves.  If
    we knew that $\rho_5$ were modular, then the computation above would
    apply and $E$ would be modular.
    
    We will find a new semistable elliptic curve $E^\prime$ such that
    $\rho_{E,5} = \rho_{E^\prime,5}$ and $\rho_{E^\prime,3}$ is
    irreducible.  Then by Case I, $E^\prime$ is modular.  Therefore
    $\rho_{E,5}= \rho_{E^\prime,5}$ does have a modular lifting and we
    will be done.
    
    We need to construct such an $E^\prime$.  Let $X$ denote the modular
    curve whose points correspond to pairs $(A, C)$ where $A$ is an
    elliptic curve and $C$ is a subgroup of $A$ isomorphic to the group
    scheme $E[5]$.  (All such curves will have mod-5 representation
    equal to $\rho_E$.)  This $X$ is genus 0, and has one rational point
    corresponding to $E$, so it has infinitely many.  Now Wiles uses a
    Hilbert Irreducibility argument to show that not all rational points
    can be images of rational points on modular curves covering $X$,
    corresponding to degenerate level 3 structure (i.e. $im(\rho_3) \neq
    GL_2(Z/3)$).  In other words, an $E^\prime$ of the type we need
    exists.  (To make sure $E^\prime$ is semistable, choose it
    5-adically close to $E$.  Then it is semistable at 5, and at other
    primes because $\rho_{E^\prime,5} = \rho_{E,5}$.)
  \end{itemize}

  \subsection{If not, then what?}

  FLT is usually broken into 2 cases. The first case assumes $(abc,n) =
  1$. The second case is the general case.

  \subsubsection{What has been proved}
   

  First Case.

  It has been proven true up to $7.568*10^{17}$ by the work of Wagstaff
  \& Tanner, Granville \& Monagan, and Coppersmith. They all used
  extensions of the Wiefrich criteria and improved upon work performed
  by Gunderson and Shanks \& Williams.

  The first case has been proven to be true for an infinite number of
  exponents by Adelman, Frey, et. al. using a generalization of the
  Sophie Germain criterion


  \noindent Second Case:

  It has been proven true up to $n = 150,000$ by Tanner \& Wagstaff. The
  work used new techniques for computing Bernoulli numbers mod $p$ and
  improved upon work of Vandiver. The work involved computing the
  irregular primes up to 150,000. FLT is true for all regular primes by
  a theorem of Kummer. In the case of irregular primes, some additional
  computations are needed. More recently, Fermat's Last Theorem has been
  proved true up to exponent 4,000,000 in the general case. The method
  used was essentially that of Wagstaff: enumerating and eliminating
  irregular primes by Bernoulli number computations. The computations
  were performed on a set of NeXT computers by Richard Crandall et al.

  Since the genus of the curve $a^n + b^n = 1$, is greater than or equal
  to 2 for $n > 3$, it follows from Mordell's theorem [proved by
  Faltings], that for any given $n$, there are at most a finite number
  of solutions.


  \subsubsection{Conjectures}

  There are many open conjectures that imply FLT. These conjectures come
  from different directions, but can be basically broken into several
  classes: (and there are interrelationships between the classes)

  \begin{enumerate}
    \item Conjectures arising from Diophantine approximation theory such
    as the ABC conjecture, the Szpiro conjecture, the Hall conjecture,
    etc.

    For an excellent survey article on these subjects see the article by
    Serge Lang in the Bulletin of the AMS, July 1990 entitled ``Old and
    new conjectured Diophantine inequalities''.

    Masser and Osterle formulated the following known as the ABC
    conjecture:

    Given $\epsilon > 0$, there exists a number $C(\epsilon)$ such that
    for any set of non-zero, relatively prime integers $a,b,c$ such that
    $a+b = c$ we have $\max( |a|, |b|, |c|) \leq C(\epsilon) N(abc)^{1 +
      \epsilon}$ where $N(x)$ is the product of the distinct primes
    dividing $x$.

    It is easy to see that it implies FLT asymptotically. The conjecture
    was motivated by a theorem, due to Mason that essentially says the
    ABC conjecture is true for polynomials.

    The ABC conjecture also implies Szpiro's conjecture [and vice-versa]
    and Hall's conjecture. These results are all generally believed to
    be true.

    There is a generalization of the ABC conjecture [by Vojta] which is
    too technical to discuss but involves heights of points on
    non-singular algebraic varieties . Vojta's conjecture also implies
    Mordell's theorem [already known to be true]. There are also a
    number of inter-twined conjectures involving heights on elliptic
    curves that are related to much of this stuff. For a more complete
    discussion, see Lang's article.

    \item Conjectures arising from the study of elliptic curves and
    modular forms. -- The Taniyama-Weil-Shimura conjecture.

    There is a very important and well known conjecture known as the
    Taniyama-Weil-Shimura conjecture that concerns elliptic curves.
    This conjecture has been shown by the work of Frey, Serre, Ribet,
    et. al. to imply FLT uniformly, not just asymptotically as with the
    ABC conj.

    The conjecture basically states that all elliptic curves can be
    parameterized in terms of modular forms.

    There is new work on the arithmetic of elliptic curves. Sha, the
    Tate-Shafarevich group on elliptic curves of rank 0 or 1. By the way
    an interesting aspect of this work is that there is a close
    connection between Sha, and some of the classical work on FLT. For
    example, there is a classical proof that uses infinite descent to
    prove FLT for $n = 4$. It can be shown that there is an elliptic
    curve associated with FLT and that for $n=4$, Sha is trivial. It can
    also be shown that in the cases where Sha is non-trivial, that
    infinite-descent arguments do not work; that in some sense 'Sha
    blocks the descent'. Somewhat more technically, Sha is an
    obstruction to the local-global principle [e.g. the Hasse-Minkowski
    theorem].

    \item Conjectures arising from some conjectured inequalities
    involving Chern classes and some other deep results/conjectures in
    arithmetic algebraic geometry.

    This results are quite deep. Contact Barry Mazur [or Serre, or
    Faltings, or Ribet, or ...]. Actually the set of people who DO
    understand this stuff is fairly small.

    The Diophantine and elliptic curve conjectures all involve deep
    properties of integers. Until these conjecture were tied to FLT, FLT
    had been regarded by most mathematicians as an isolated problem; a
    curiosity. Now it can be seen that it follows from some deep and
    fundamental properties of the integers. [not yet proven but
    generally believed].

  \end{enumerate}
}

\subsection{Related Conjectures}

A related conjecture from Euler

\[ x^n + y^n + z^n = c^n \mbox{ has no solution if n is} \geq 4 \]

Noam Elkies gave a counterexample, namely $2682440^4 + 15365639^4 +
18796760^4 = 20615673^4$. Subsequently, Roger Frye found the absolutely
smallest solution by (more or less) brute force: it is $95800^4 +
217519^4 + 414560^4 = 422481^4$.  ``Several years'', Math. Comp. 51
(1988) 825-835.
 

This synopsis is quite brief. A full survey would run too many pages.

\Ref

\article{[1] J.P.Butler, R.E.Crandall,\& R.W.Sompolski,} {Irregular
  Primes to One Million.}  {Math. Comp.,} {59 (October 1992)
  pp. 717-722.}

\book{Fermat's Last Theorem, A Genetic Introduction to Algebraic Number
  Theory.} {H.M. Edwards.} {Springer Verlag, New York, 1977.}

\book{Thirteen Lectures on Fermat's Last Theorem.} {P. Ribenboim.}
{Springer Verlag, New York, 1979.}

\book{Number Theory Related to Fermat's Last Theorem.} {Neal Koblitz,
  editor.} {Birkh\"auser Boston, Inc., 1982, ISBN 3-7643-3104-6}

\subsection{Did Fermat prove this theorem?}

No he did not. Fermat claimed to have found a proof of the theorem at an
early stage in his career. Much later he spent time and effort proving
the cases $n=4$ and $n=5$. Had he had a proof to his theorem earlier,
there would have been no need for him to study specific cases.

Fermat may have had one of the following ``proofs'' in mind when he
wrote his famous comment.

\begin{itemize}
  \item Fermat discovered and applied the method of infinite descent,
  which, in particular can be used to prove FLT for $n=4$.  This method
  can actually be used to prove a stronger statement than FLT for $n=4$,
  viz, $x^4 + y^4 = z^2$ has no non-trivial integer solutions.  It is
  possible and even likely that he had an incorrect proof of FLT using
  this method when he wrote the famous ``theorem''.
  \item He had a wrong proof in mind. The following proof, proposed
  first by Lame' was thought to be correct, until Liouville pointed out
  the flaw, and by Kummer which latter became and expert in the field.
  It is based on the {\em incorrect} assumption that prime decomposition
  is unique in all domains.

  The incorrect proof goes something like this:

  We only need to consider prime exponents (this is true).  So consider
  $x^p + y^p = z^p$.  Let $r$ be a primitive $p$-th root of unity
  (complex number)

  Then the equation is the same as:

  \[(x+y)(x+ry)(x+r^2y)...(x+r^{p-1}y) = z^p\]

  Now consider the ring of the form:

  \[a_1 + a_2 r + a_3 r^2 + ... + a_{p-1} r^{p-1}\]

  where each $a_i$ is an integer

  Now {\bf if} this ring is a unique factorization ring (UFR), then it
  is true that each of the above factors is relatively prime.

  From this it can be proven that each factor is a $p$th power from
  which FLT follows.  This is usually done by considering two cases, the
  first where $p$ divides none of $x$, $y$, $z$; the second where $p$
  divides some of $x$, $y$, $z$.  For the first case, if $x+yr=u*t^p$,
  where $u$ is a unit in $Z[r]$ and $t$ is in $Z[r]$, it follows that
  $x=y (mod p)$. Writing the original equation as $x^p + (-z)^p =
  (-y)^p$, it follows in a similar fashion that $x = -z (mod p)$.  Thus
  $2*x^p = x^p + y^p = z^p = -x^p (mod p)$ which implies $3*x^p = 0 (mod
  p)$ and from there $p$ divides one of $x$ or $3|x$. But $p>3$ and $p$
  does not divides $x$; contradiction.  The second case is harder.


  The problem is that the above ring is {\bf not} an UFR in general.

\end{itemize}

Another argument for the belief that Fermat had no proof---and,
furthermore, that he {\bf knew} that he had no proof---is that the only
place he ever mentioned the result was in that marginal comment in
Bachet's Diophantus. If he really thought he had a proof, he would have
announced the result publicly, or challenged some English mathematician
to prove it. It is likely that he found the flaw in his own proof before
he had a chance to announce the result, and never bothered to erase the
marginal comment because it never occurred to him that anyone would see
it there.

Some other famous mathematicians have speculated on this question.
Andre Weil, writes:
\begin{quote}
  Only on one ill-fated occasion did Fermat ever mention a curve of
  higher genus $x^n+y^n=z^n$, and then hardly remains any doubt that
  this was due to some misapprehension on his part [$\ldots$] for a
  brief moment perhaps [$\ldots$] he must have deluded himself into
  thinking he had the principle of a general proof.
\end{quote}

Winfried Scharlau and Hans Opolka report:
\begin{quote}
  Whether Fermat knew a proof or not has been the subject of many
  speculations.  The truth seems obvious $\ldots$ [Fermat's marginal
  note] was made at the time of his first letters concerning number
  theory [1637]$\ldots$ as far as we know he never repeated his general
  remark, but repeatedly made the statement for the cases $n=3$ and $4$
  and posed these cases as problems to his correspondents [$\ldots$] he
  formulated the case $n=3$ in a letter to Carcavi in 1659 [$\ldots$]
  All these facts indicate that Fermat quickly became aware of the
  incompleteness of the [general] ``proof'' of 1637.  Of course, there
  was no reason for a public retraction of his privately made
  conjecture.
\end{quote}

However it is important to keep in mind that Fermat's ``proof'' predates
the Publish or Perish period of scientific research in which we are
still living.

\Ref

\book{From Fermat to Minkowski: lectures on the theory of numbers and
  its historical development.}{ Winfried Scharlau, Hans Opolka.}  {New
  York, Springer, 1985.}

\book{Basic Number Theory.}{Andre Weil.}{Berlin, Springer, 1967}
