\section{Why a list of Frequently Asked Questions?}

\emph{The Net}, as users call the Internet, and specially
\emph{newsgroups}, (i.e., Usenet) created a demand of knowledge without
parallel since the invention of the printing press.  Surprisingly, the
type of knowledge demanded from and by the Usenet community had, in most
cases, little in common---both in structure and content---with that of
printed in current publications. This defined Usenet as more of an
alternative to books rather than a replacement thereof.\footnote{It
  could be argued that books fulfill their mandate and purpose to
  everybody's satisfaction.  Thus, even though the Net could, in
  principle, replace the need for books, people choose not to do
  so. Instead it's domain is defined, by its very nature to be disjoint
  from books.}

In the Net, questions posed are, more often than not, at the level of an
amateur practitioner---even in cases where the question was posed by a
professional in the field. Similarly, the quality of the answers varies
greatly, ranging from the incorrect or disrespectful, to summaries of
the state of the art in the topic in question.

Other characteristics of communication on the Net are simply inherited
from restrictions of the medium. The unit of knowledge is a screenful
worth of text (a scrit, from screen and bit). Articles exceeding that
limit are usually disregarded.

The lack of memory of the medium generates a repetition of topics, much
to the chagrin of old time citizens of the Net.  Frequently asked
questions (FAQ) lists palliate some of these deficiencies by providing a
record of relevant information while at the same time never being
outdated.\footnote{Or, at least, the editors would like to think so!}

Thus, typically a list of frequently asked questions is ``posted'' at
least once a month, and updated at least as frequently. And, in what
must be a first for an information based product, FAQ lists ``expire''
on a given date, very much like any other perishable item.


\section{Frequently Asked Questions in Mathematics?}

If I had to describe the contents of the FAQ in Mathematics in a single
sentence, I would call it \emph{mathematical gossip} or perhaps
\emph{non-trivial mathematical trivia}.

The FAQ list is a compilation of knowledge of interest to most
professional and amateur mathematicians, ranging from advanced topics
such as Wiles' proof to Fermat's Last Theorem to the list of Fields
Medal winners.

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "math-faq"
%%% End: 
